---
title: "Prediction of Weight Lifting Excercises"
author: "Dr. Ivan Bleskov"
date: "2/5/2017"
output:
  html_document: default
  pdf_document: default
---
```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Introduction
### Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity. One thing that people
regularly do is quantify how much of a particular activity they do, but they 
rarely quantify how well they do it. In this project, our goal is to use data 
from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 
They were asked to perform barbell lifts correctly and incorrectly in 5 
different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har.

### Data
The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

### Goal
The goal of the work is to predict the manner (20 different test cases) in which
the exercise is performed. This is the “classe” variable in the training set.

## Data preparation
### Loading
Load the needed packages and datasets.
```{r}
set.seed(137)
require(dplyr)
require(data.table)
require(caret)

url_tr <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_te <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

tr <- fread(url_tr, na.strings = c("NA", "", "#DIV/0!"))
te <- fread(url_te, na.strings = c("NA", "", "#DIV/0!"))

dim(tr); dim(te)
```

### Cleaning
We keep only those columns (predictors) which (i) contain no NA's and (ii) are 
related to data from accelerometers on the belt, forearm, arm, and dumbell.
```{r}
ll <- lapply(
    list(tr, te),
    # The function selects only columns with proper names and without NA's
    function(df) {
        df <- df %>% 
            select( 
                which( 
                    # cols without NA's
                    (colSums(is.na(.)) == 0) &
                    # cols containing relevant names
                    (grepl("dumbbell|forearm|_arm|belt", names(.))) ) )
        return(df) } )
# In the "cleaned" train dataset we, of course, need an outcome column "classe"
tr <- cbind("classe" = as.factor(tr[, classe]),
            ll[[1]])
# In the "cleaned" test dataset we need only predictors
te <- ll[[2]]
# memory emptying
ll <- NULL
```

### Splitting
We split the train dataset into sub-train (```tr_tr```) and validation 
(```tr_vld```) datasets.
```{r}
tt <- createDataPartition(tr$classe, p = .7, list = FALSE)
tr_tr  <- tr[ tt, ]
tr_vld <- tr[-tt, ]
# splitting check
dim(tr_tr)[1] + dim(tr_vld)[1] == nrow(tr)
```

Check for predictors with no (near zero) variability which are useless in a
prediction model constuction:
```{r}
unique(nearZeroVar(tr_tr, saveMetrics = TRUE)$`nzv`)
```

There are no such predictors.

## Train and prediction on a validation set
We choose the random forest prediction algorithm with 5-fold cross 
validation. Since this is rather powerful approach, we expect high prediction
accuracy.
```{r}
fit <- train(classe ~ .,
             data = tr_tr,
             method = "rf",
             trControl = trainControl(method = "cv",
                                      number = 5))
fit$finalModel
```

OOB error is small ($< 1\%$).
The training took quite some time, therefore it's better to save the model for 
the further use.
```{r, eval = FALSE}
save(fit, file = "fit.RData")
```

We now use the model to predict on the validation set:
```{r}
pred_tr <- predict(fit, newdata = tr_vld)
confusionMatrix(tr_vld$classe, pred_tr)
```

The accuracy is extremely high ($.998$). Obviously, this model is good enough 
and we don't need to train other models on the other validation sets.

## Prediction on a test set
We now use our model to predict the manner of excercise performance on the 
unknown test set.
```{r}
pred_te <- predict(fit, newdata = te)
```

Here are the predicted manners
```{r}
pred_te
```